GPU-friendly Stroke Expansion

### 2024-06-14 paper.tex
Minor edits and figure tweaks
A bit more detail on the subdivision density integral approximation. Remove mmark from timings graph. Minor edits in response to review feedback.

Stroke rendering is an essential part of the standard vector graphics imaging model, and is thus widely used. The standard representation of paths is a sequence of cubic Bézier segments. While there are a number of published techniques for GPU rendering of filled paths, stroke rendering is more challenging to parallelize, largely because path segments cannot be processed independently of each other; rendering of the \emph{joins} between path segments depends on both adjoining segments, and the endpoints of paths (when not closed) are rendered with \emph{caps}. Joins and caps can have multiple styles, including different styles within the same document.

A typical vector graphics document may have thousands of stroked paths. A map or CAD drawing may have tens of thousands of paths, not unusually a million path segments. At this scale, the computational cost of stroke rendering on CPU may then result in loss of interactive rendering performance. Thus, we desire to improve performance by offloading stroke rendering to the GPU. To realize performance gains, the algorithm needs to be GPU-friendly, meaning among other things that it exploits the available parallelism, is work efficient, and avoids unnecessarily complex (and potentially divergent) control flow. In addition, it must remain numerically robust even when computed with 32 bit floating point numbers.

There are a number of strategies for rendering strokes. Among them are \emph{local} techniques that break down the stroke into individual closed primitives, distance field techniques (or similar techniques based on point inclusion). In this paper, we focus on \emph{stroke expansion,} the generation of an outline that, when filled, represents the rendered stroke. Implementing stroke rendering using stroke expansion enables a unified approach to path rendering downstream, rendering both filled and stroked paths. Among other benefits, such an approach avoids a large number of draw calls when stroked and filled paths are finely interleaved, as is often the case.

Correctness is another challenging aspect of stroke rendering. We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins. Weak correctness, by contrast, only requires the parallel curves (also known as ``offset curves'') of path segments, combined with caps and the outer contours of joins. Both \citet{Nehab2020} and \citet{Kilgard2020} provide useful definitions of strongly correct stroke rendering, and Nehab in particular describes how to achieve it in stroke expansion. Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, then addition \emph{evolute} segments need to be added, in addition to the parallel curves required by weak correctness, and also the inner contours of joins are required in general. The technique described in this paper is intended for strongly correct stroke rendering; it produces outlines that render the same shape as \citet{Nehab2020}.

Correct stroke rendering also relies on accurate approximation of the parallel curves of the input paths. Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula[\citenum{Farouki1990}]. Thus, all practical stroke rendering techniques employ an approximation, with error tolerance as a tunable parameter, typically representing the maximum allowable Fréchet distance between the true curve and its approximation. A typical value is 0.25 device pixel, as that is close to the threshold of visibility. A correct algorithm will produce an approximate curve within the error tolerance, but an efficient algorithm will not subdivide significantly more than is necessary to meet the error bound. Some specifications for vector graphics, including PDF[\citenum{PDF2008}], specify the error tolerance explicitly.

The subproblem of approximating parallel curves has spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU. An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library. For producing curved outlines, the Tiller and Hanson[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers. The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}. A variant is also used in Skia[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.

@@

Our central approach bounding the error tolerance is \emph{error metrics,} which predict the Fréchet distance error. The error metrics presented in this paper are both straightforward to compute, and also \emph{invertible}, which avoids excessive subdivision and also increases the amount of available parallelism to exploit. Previous approaches to error measurement generally follow a ``cut then measure'' approach, where the approximate curve is generated, then error measured by sampling. Such sampling is often time consuming, involving iteration, and even then risks underestimating the error due to inadequate sampling.

This paper presents a comprehensive solution to stroke expansion, well suited to GPU implementation. It can produce both flattened polylines or an approximation consisting of circular arcs. The best choice depends on the capabilities of the path rendering mechanism following stroke expansion, but as we show, outlines consisting of circular arcs have many fewer segments and are correspondingly faster to produce. The algorithm is based on Euler spiral segments as an intermediate representation, with an iterative algorithm based on a straightforward error metric for conversion from cubic Béziers. We have also devised a compact binary encoding of paths, suitable for fully parallel computation of stroke outlines, while requiring minimal CPU-side processing. Our algorithm has been implemented in GPU compute shaders, integrated in a full rendering engine for vector graphics, and shows a dramatic speedup over CPU methods.

@@

The distance between a circular arc segment of length $s$ and its chord, with angle between arc and chord of $\theta$ (see Figure~\ref{fig:arc_chord}), is exactly $(1-\cos \theta)\frac{s}{2\theta}$. The curvature is $\kappa = \frac{2\theta}{s}$ (equivalently, $\theta = \frac{\kappa s}{2}$), and this remains constant even as the arc is subdivided. Rewriting, $d = (1-\cos\frac{\kappa s}{2})\frac{1}{\kappa}$. From this, we can derive a precise, invertible error metric. Subdividing the arc in to $n$ segments, the distance error for each segment is $\frac{1}{\kappa}(1-\cos \frac{\kappa s}{2n})$. Solving for n, we get:

@@

We are of course concerned with the flattening of more general curves (ultimately the parallel curve of a cubic Bézier), not simply circular arcs, so the curvature is not constant. An obvious approach to try would be using the maximum error. This would be conservative with respect to error tolerance, but also generate too much subdivision when the curve has a small region of high curvature. In the limit, the curve can contain a cusp of infinite curvature, which would entail infinite subdivision. Similarly, sampling the curvature at a single point can either underestimate or overestimate the error. The ideal approach would be a form of average curvature that is tractable to compute, and accurately predicts the flattening error. To that end, we propose the following error metric to estimate the maximum distance between an arbitrary curve of length $\hat{s}$ and its chord.

@@

\begin{figure}
    \includegraphics[scale=0.3]{es_flatten}
    \caption{The flattened approximation to an Euler spiral using line segments is achieved by choosing subdivision points using a simple power law formula with an exponent of 2/3.}
    \Description{An Euler spiral along with its flattened representation. The flattened curve has a consistent error, and the subdivision points are spaced according to a}
    \label{fig:es_flatten}
\end{figure}

@@

It is common to approximate cubic Béziers to some intermediate curve format more conducive to offsetting and flattening. A number of published solutions (\citet{Yzerman2020}, \citet{Nehab2020}) use quadratic Béziers, as it is well suited for computation of parallel curves. Even so, this curve has some disadvantages. In particular, it cannot model an inflection point, so the source curve must be subdivided at inflection points.

@@

The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length, of the form $\kappa(s) = c_0{s - s_0}^{-1/2} + c_1$.

@@

The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segments, offset by the half linewidth. The joins and caps are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky. Analytically, it is a tenth order algebraic curve, which is not particularly feasible to compute directly.

@@

Using a subpixel Fréchet distance bound guarantees that the rendered image does not deviate visibly from the exact rendering. Another choice would be uniform steps in tangent angle, as chosen by polar stroking[\citenum{Kilgard2020}]. However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering.

The limitation of the angle step error metric is shown in Figure~\ref{fig:angle_err}. The top row shows the use of a distance-based error metric, as is used in our approach, which is visually consistent at varying curvature (in practice, a tolerance value of 0.25 of a pixel is below the threshold of perceptibility). The bottom row shows a consistent angle step, as implemented in polar stroking, but has excessive distance error at low curvature, and excessive subdivision at high curvature. It should be noted, to avoid the undershoot at low curvature, both the Skia[\citenum{Skia}] and Rive[\citenum{Rive}] renderers use a hybrid of the Wang and polar stroking error metrics.

@@

The subdivision density integral (Section~\ref{subsection:subdiv-density-int}) is fairly straightforward to compute in the forward direction, but not invertible using a straightforward closed-form equation. Numerical techniques are possible, but require multiple iterations to achieve sufficient accuracy, so are slower. In this subsection, we present a straightforward and accurate approximation, constructed piecewise from easily invertible functions. If higher flattening quality is desired at the expense of slower computation, this approximation can be used to determine a good initial value for numeric techniques; two iterations of Newton solving are enough to refine this guess to within 32-bit floating point accuracy.

The approximation is given as follows:

@@

The exact integral and the approximation given above are shown in Figure~\ref{fig:espc}. Visually, it is clear that the agreement is close, and in numerical testing the worst case discrepancy between approximate and exact results is approximately 6\%.

@@

See Figure~\ref{fig:g_comparison} for a comparison between flattening and approximation with arc segments. The arc segment version has many fewer segments at the same tolerance, while preserving very high visual quality.

@@

In the principled, correct specification for stroking[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width. When it does, additional segments must be drawn, including evolutes of the original curve. In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques. By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve[\citenum{Yoshida2012}].

@@

Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type that is more tractable for evaluating parallel curve. Computing parallel curves directly on cubic Bézier curve segments is not very tractable. In particular, the widely cited Tiller-Hanson algorithm[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.

@@

A typical pattern for converting from one curve type to another is \emph{adaptive subdivision.} An approximate curve is found in the parameter space of the target curve family. The error of the approximation is measured. If the error exceeds the specified tolerance, the curve is subdivided (typically at $t = 0.5$), otherwise the approximation is accepted. Subdivisions are also indicated at special points; for example, since quadratic Béziers cannot represent inflection points, and geometric Hermite interpolation is numerically unstable if the input curve is not convex, lowering to quadratic Béziers also requires calculation of inflection points and subdividing there. A good example of this pattern is \citet{Nehab2020}. One advantage of Euler spirals over quadratic Béziers is that they can represent inflection points just fine, so it is not necessary to compute those for additional subdivision.

@@

Our approach is different. In short, we perform a straightforward analytical computation to accurately estimate the error. Our approach to the error metric has two major facets. First, we obtain a cubic Bézier which is a very good fit to the Euler spiral, then we estimate the distance between that and the source cubic. Due to the triangle inequality, the sum of these is a conservative estimate of the true Fréchet distance between the cubic and the Euler spiral.

@@

Given tangent angles relative to the chord, finding the Euler spiral segment that minimizes total curvature variation is a form of geometric Hermite interpolation. There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent[\citenum{Kimia2003}], bisection[\citenum{Walton2009}], or Newton iteration[\citenum{Connor2014}]. A more direct approach is to approximate the function as a reasonably low-order polynomial in terms of the endpoint angles. Our approach is to use a 7th order polynomial, which is more precise than the 3rd order polynomial proposed in \citet{Reif2021}, at only modest increased cost. The exact coefficients used are presented in Appendix \ref{appendix:gh}.

@@

The entire recursion stack for adaptive subdivision can be represented in two words: $dt$ and $t0\_u$ (the latter being an unsigned integer). The range is $dt \cdot t0\_u .. dt \cdot (t0\_u + 1)$. Initial values of $1$ and $0$ represent the range $(0, 1)$. Pushing the stack, subdividing the range in half, is represented by halving $dt$ and doubling $t0\_u$, which leaves the start of the range invariant but its size halved. After accepting an approximation, the next range is determined by incrementing $t0\_u$, then repeatedly popping the stack by doubling $dt$ and halving $t0\_u$, as long as the latter is odd. This can in fact be achieved without iteration by using the ``count trailing zeros'' intrinsic.

@@

A Bézier curve is expressed in parametric form, and the derivative with respect to the parameter can be zero or nearly so, causing serious numerical problems for many stroke expansion algorithms. In the limit, as the Bézier curve describes a semi-cubical parabola, the curvature can become infinite.

@@

\begin{figure*}
    \includegraphics[scale=0.19]{stroke_stills}
    \caption{Renderings of stroked cubic Béziers exhibiting strong and weak correctness. The outlines generated by our GPU stroking compute shader are visualized in green and blue. Segments with a winding direction that points \emph{up} are colored blue while those pointing \emph{down} are colored green. The curves were adapted to from the \emph{stills} dataset by \citet{Nehab2020} and the Skia test corpus.}
    \Description{Selection of stroked cubic Béziers exhibiting strong and weak correctness}
    \label{fig:stroke_stills}
\end{figure*}

@@

One of our artifacts is a GPU compute shader that satisfies these criteria. We coupled this with a simple and efficient data layout for parallel processing. Notably, our implementation is fully data-parallel on input path segments and does not require any computationally expensive curve evaluation on the CPU. We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU[\citenum{WebGPU}]. As such, the ideas presented in this section are portable to various GPU platforms.

@@

An SVG path consists of a sequence of instructions (or "verbs"). The \emph{lineto} and \emph{curveto} instructions denote an individual \emph{path segment} consisting of either a straight line or a cubic Bézier. A given subsequence of path segments can be bounded by a \emph{moveto} and/or \emph{closepath} instruction that marks the beginning and end of a \emph{subpath}. A subpath must form a closed contour if painted as a fill. When painted as a stroke, each subpath must begin and end with a cap according to the desired cap style. In addition, each path segment in a stroked subpath must be connected to adjacent path segments of the same subpath with a join. For any given path, the cap and join styles do not vary across subpaths or individual path segments.

@@

\begin{algorithm}
  $tid \gets$ global invocation ID\;
  $(style, segment) \gets readScene(tid)$\;
  \eIf{$style$ is $fill$}{
      \tcp{Output source curve}
      flatten\_cubic(segment, $0$)\;
  }{
      \tcp{Handle stroked segment}
      \eIf{$segment$ is \emph{stroke cap marker}}{
          \eIf{path is open}{output start cap\;}{output nothing\;}
      }{
          \tcp{Output both parallel curves}
          flatten\_cubic(segment,  style.offset)\;
          flatten\_cubic(segment, -style.offset)\;

@@

Each thread outputs a polyline that fits a subset of the rendered subpath's outer contour, represented by the path segment (see Figure \ref{fig:stroke_threads}). For a filled primitive, a thread only outputs the flattened approximation of the path segment. For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{halfwidth}, and b) the joins that connect the path segment to the next adjacent segment, and c) evolute patches in regions of high curvature. We also store metadata in our input encoding that marks the position of a path segment within the containing subpath. We use this information to determine whether to output an end cap instead of a join. We encode one additional segment designated as the \emph{stroke cap marker} for every subpath. The thread assigned to the marker segment only outputs a start cap, which completes the stroke outline.

@@

We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, and two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}. We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}[\citenum{wgpu}] framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).

@@

    \item Dashing stroked paths on the GPU was left out of scope for this paper. Given its parameterization by arc length, we believe the Euler spiral representation is highly suitable for dashing.
\end{itemize}

>>>>

Stroke rendering is an essential and widely part of the standard vector graphics imaging model. The standard representation of paths is a sequence of cubic Bézier segments. While there are a number of published techniques for GPU rendering of filled paths, stroke rendering is more challenging to parallelize, largely because path segments cannot be processed independently of each other; rendering of the \emph{joins} between path segments depends on both adjoining segments, and the endpoints of paths (when not closed) are rendered with \emph{caps}. Joins and caps can be rendered with a variety of styles within the same document. The style of stroke rendering also optionally includes \emph{dashing}, which applies a pattern of dashes and gaps to the outline of the shape, as illustrated in Figure~\ref{fig:stroke_styles}.

\begin{figure}
    \vspace{15pt}
    \includegraphics[scale=0.4]{stroke_styles}
    \caption{Illustration of stroke styles applied to a source Bézier path (shown in black).}
    \Description{Illustration of stroke styles applied to a source Bézier path (shown in black).}
    \label{fig:stroke_styles}
\end{figure}

A typical vector graphics document may have thousands of stroked paths. A map or CAD drawing may have tens of thousands of paths, possibly consisting of millions of path segments. At this scale, the computational cost of stroke rendering on a CPU may result in loss of interactive rendering performance. We believe the runtime performance can be improved significantly by offloading stroke rendering to a GPU. This requires that the algorithm be GPU-friendly: that it exploit the available parallelism, be work efficient, and avoid unnecessarily complex and divergent control flow. It must also remain numerically robust when computed with 32-bit floating pointing numbers.

There are a number of strategies for rendering strokes. Among them are \emph{local} techniques that break down the stroke into individual closed primitives, distance field techniques (or similar techniques based on point inclusion). In this paper, we focus on \emph{stroke expansion,} the generation of an outline that, when filled, represents the rendered stroke. Implementing stroke rendering using stroke expansion enables a unified approach to path rendering both filled and stroked paths downstream. Among other benefits, such an approach avoids a large number of draw calls when stroked and filled paths are finely interleaved, as is often the case.

Correctness is another challenging aspect of stroke rendering. We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins. Weak correctness, by contrast, only requires the parallel curves (also known as ``offset curves'') of path segments, combined with caps and the outer contours of joins. Both \citet{Nehab2020} and \citet{Kilgard2020} provide useful definitions of strongly correct stroke rendering, and Nehab in particular describes how to achieve it in stroke expansion. Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, \emph{evolute} segments and the \emph{inner contours of joins} are required in addition to the parallel curves. The technique described in this paper produces strongly correct outlines according to this definition. Examples of weakly and strongly correct rendering are shown in Figure~\ref{fig:weak_strong}.

\begin{figure}
    \includegraphics[scale=0.5]{weak_strong}
    \caption{Example of weakly correct (top) and strongly correct (bottom) stroke rendering. The left example has a curvature cusp and requires evolutes to render correctly, while the right requires the inner join.}
    \Description{Two examples of strokes. The top left is missing an evolute section, making it appear that a half-disc has been carved out. The top right is missing inner joins, looking like spokes rather than a connected path. The bottom two examples are rendered with the full strokes.}
    \label{fig:weak_strong}
\end{figure}

Correct stroke rendering also relies on accurate approximation of the parallel curves of the input paths. Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula~[\citenum{Farouki1990}]. Thus, all practical stroke rendering techniques employ an approximation, with error tolerance as a tunable parameter, typically representing the maximum allowable Fréchet distance between the true curve and its approximation. A typical value is 0.25 device pixel, as that is close to the threshold of visibility. A correct algorithm will produce an approximate curve within the error tolerance, but an efficient algorithm will not subdivide significantly more than is necessary to meet the error bound. Some specifications for vector graphics, including PDF~[\citenum{PDF2008}], specify the error tolerance explicitly.

The subproblem of approximating parallel curves has spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU. An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library. For producing curved outlines, the Tiller and Hanson~[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers. The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}. A variant is also used in Skia~[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.

@@

Our central approach bounding the error tolerance is \emph{error metrics,} which predict the Fréchet distance error. The error metrics presented in this paper are both straightforward to compute, and also \emph{invertible}, which avoids excessive subdivision and also increases the amount of available parallelism to exploit. Previous approaches to error measurement generally follow a ``cut then measure'' approach, where the approximate curve is generated, then the error measured by sampling. Such sampling is often time consuming, involving iteration, and even then risks underestimating the error due to inadequate sampling.

Previous work, particularly \citet{Kilgard2020}, has explained how to break the input path into dash segments of the specified length on GPU using a prefix sum of segment arc lengths. There are implementation tradeoffs to GPU-based dash processing, including potentially a larger number of dispatch stages. We anticipate that the potential performance benefit of implementing dashing on a GPU depends on the scene and other factors in the overall rendering pipeline. While we would expect further performance improvement from GPU-based dashing, we did not fully explore the solution space and we were able to achieve acceptable performance with dash processing on CPU.

This paper presents a comprehensive solution to the core problem of stroke expansion, well suited to GPU implementation. It can produce both flattened polylines or an approximation consisting of circular arcs. The best choice depends on the capabilities of the path rendering mechanism following stroke expansion, but as we show, outlines consisting of circular arcs have many fewer segments and are correspondingly faster to produce. The algorithm is based on Euler spiral segments as an intermediate representation, with an iterative algorithm based on a straightforward error metric for conversion from cubic Béziers. We have also devised a compact binary encoding of paths, suitable for fully parallel computation of stroke outlines, while requiring minimal CPU-side processing. Our algorithm has been implemented in GPU compute shaders, integrated in a full rendering engine for vector graphics, and shows a dramatic speedup over CPU methods.

@@

The distance between a circular arc segment of length $s$ and its chord, with angle between arc and chord of $\theta$ (see Figure~\ref{fig:arc_chord}), is exactly $(1-\cos \theta)\frac{s}{2\theta}$. The curvature is $\kappa = \frac{2\theta}{s}$ (equivalently, $\theta = \frac{\kappa s}{2}$), and this remains constant even as the arc is subdivided. Rewriting, $d = (1-\cos\frac{\kappa s}{2})\frac{1}{\kappa}$. From this, we can derive a precise, invertible error metric. Subdividing the arc into $n$ segments, the distance error for each segment is $\frac{1}{\kappa}(1-\cos \frac{\kappa s}{2n})$. Solving for $n$, we get:

@@

We are of course concerned with the flattening of more general curves (ultimately the parallel curve of a cubic Bézier), not simply circular arcs, so the curvature is not constant. An obvious approach to try would be to use the maximum error. This would be conservative with respect to error tolerance, but also generate too much subdivision when the curve has a small region of high curvature. In the limit, the curve can contain a cusp of infinite curvature, which would entail infinite subdivision. Similarly, sampling the curvature at a single point can either underestimate or overestimate the error. The ideal approach would be a form of average curvature that is tractable to compute, and accurately predicts the flattening error. To that end, we propose the following error metric to estimate the maximum distance between an arbitrary curve of length $\hat{s}$ and its chord.

@@

\begin{figure}
    \includegraphics[scale=0.3]{es_flatten}
    \caption{The flattened approximation to an Euler spiral using line segments is achieved by choosing subdivision points using a simple power law formula with an exponent of 2/3.}
    \Description{An Euler spiral along with its flattened representation. The flattened curve has a consistent error, and the subdivision points are spaced according to a power law.}
    \label{fig:es_flatten}
\end{figure}

@@

It is common to approximate cubic Béziers to some intermediate curve format more conducive to offsetting and flattening. A number of published solutions (\citet{Yzerman2020}, \citet{Nehab2020}) use quadratic Béziers, as it is well suited for computation of parallel curves. Even so, this curve has some disadvantages. For one, it cannot model an inflection point, so the source curve must be subdivided at inflection points.

@@

The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length, of the form $\kappa(s) = c_0{s - s_0}^{-1/2} + c_1$. Euler spirals as an intermediate curve representation for computing parallel curves was proposed in \citet{Levien2021}, which also expands on the Wieleitner reference (including an English translation) and Cesàro equation.

@@

The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segments, offset by the half linewidth. The joins and caps are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky~[\citenum{Kilgard2020a}]. Analytically, it is a tenth order algebraic curve, which is not particularly feasible to compute directly.

@@

Using a subpixel Fréchet distance bound guarantees that the rendered image does not deviate visibly from the exact rendering. Another choice would be uniform steps in tangent angle, as chosen by polar stroking~[\citenum{Kilgard2020}]. However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering.

The limitation of the angle step error metric is shown in Figure~\ref{fig:angle_err}. The top row shows the use of a distance-based error metric, as is used in our approach, which is visually consistent at varying curvature (in practice, a tolerance value of 0.25 of a pixel is below the threshold of perceptibility). The bottom row shows a consistent angle step, as implemented in polar stroking, but has excessive distance error at low curvature, and excessive subdivision at high curvature. It should be noted, to avoid the undershoot at low curvature, both the Skia~[\citenum{Skia}] and Rive~[\citenum{Rive}] renderers use a hybrid of the Wang and polar stroking error metrics.

@@

The subdivision density integral (Section~\ref{subsection:subdiv-density-int}) is fairly straightforward to compute in the forward direction, but not invertible using a straightforward closed-form equation. Numerical techniques are possible, but require multiple iterations to achieve sufficient accuracy, so are slower. In this subsection, we present a straightforward and accurate approximation, constructed piecewise from easily invertible functions. This approximation was found empirically, by interactively tuning candiadate approximation functions in the Desmos graphing calculator. The exact integral and the approximation are shown in Figure~\ref{fig:espc}. Visually, it is clear that the agreement is close. We also built a test suite to exhaustively test the subdivision count using a property testing approach, finding that the worst case discrepancy between approximate and exact results is 6\%. If higher flattening quality is desired at the expense of slower computation, this approximation could be used to determine a good initial value for numeric techniques; two iterations of Newton solving are enough to refine this guess to within 32-bit floating point accuracy.

The approximation is as follows:

@@

@@

See Figure~\ref{fig:g_comparison} for a comparison between flattening to a polyline and approximation with arc segments. The arc segment version has many fewer segments at the same tolerance, while preserving very high visual quality.

@@

In the principled, correct specification for stroking~[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half-width. When it does, additional segments must be drawn, including evolutes of the original curve. In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques. By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve~[\citenum{Yoshida2012}].

@@

Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type that is more tractable for evaluating parallel curve. Computing parallel curves directly on cubic Bézier curve segments is not very tractable. In particular, the widely cited Tiller-Hanson algorithm~[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.

@@

A typical pattern for converting from one curve type to another is \emph{adaptive subdivision.} An approximate curve is found in the parameter space of the target curve family. The error of the approximation is measured. If the error exceeds the specified tolerance, the curve is subdivided (typically at $t = 0.5$), otherwise the approximation is accepted. Subdivisions are also indicated at special points; for example, since quadratic Béziers cannot represent inflection points, and geometric Hermite interpolation is numerically unstable if the input curve is not convex, lowering to quadratic Béziers also requires calculation of inflection points and subdividing there. A good example of this pattern is \citet{Nehab2020}. One advantage of Euler spirals over quadratic Béziers is that they can represent inflection points just fine, so it is not necessary to solve for the inflection points, or do additional subdivision. Avoiding these helps make the algorithm a bit more GPU-friendly.

@@

Our approach is different. In short, we perform a straightforward analytical computation to accurately estimate the error. Our approach to the error metric has two major facets. First, we obtain a secondary cubic Bézier which is a very good fit to the Euler spiral, then we estimate the distance between that and the source cubic. Due to the triangle inequality, the sum of these is a conservative estimate of the true Fréchet distance between the cubic and the Euler spiral.

@@

Given tangent angles relative to the chord, finding the Euler spiral segment that minimizes total curvature variation is a form of geometric Hermite interpolation. There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent~[\citenum{Kimia2003}], bisection~[\citenum{Walton2009}], or Newton iteration~[\citenum{Connor2014}]. A more direct approach is to approximate the function as a reasonably low-order polynomial in terms of the endpoint angles. Our approach is to use a 7th order polynomial, which is more precise than the 3rd order polynomial proposed in \citet{Reif2021}, at only modest increased cost. The exact coefficients used are presented in Appendix \ref{appendix:gh}.

@@

The entire recursion stack for adaptive subdivision can be represented in two words: $dt$ and $t0\_u$ (the latter being an unsigned integer). The range is $dt \cdot t0\_u .. dt \cdot (t0\_u + 1)$. Initial values of $1$ and $0$ represent the range $(0, 1)$. Pushing the stack, subdividing the range in half, is represented by halving $dt$ and doubling $t0\_u$, which leaves the start of the range invariant but its size halved. After accepting an approximation, the next range is determined by incrementing $t0\_u$. At this point, the decision of whether to pop the stack or continue is dependent on whether $t0\_u$ is even or odd, respectively. If even, popping the stack is accomplished by doubling $dt$ and halving $t0\_u$, and this is repeated until $t0\_u$ becomes odd. The repeated stack popping can be accomplished without iteraction by using the ``count trailing zeros'' intrinsic to determine the number of stack pops $n$, then multiplying $dt$ by $2^n$, and shifting $t0\_u$ right by $n$ bits.

@@

A Bézier curve is expressed in parametric form, and its derivative with respect to the parameter can be zero or nearly so, causing serious numerical problems for many stroke expansion algorithms. In the limit, as the Bézier curve describes a semi-cubical parabola, the curvature can become infinite.

@@

\begin{figure*}
    \includegraphics[scale=0.19]{stroke_stills}
    \caption{Stroked cubic Béziers exhibiting strong and weak correctness, rendered by our implementation. The contents of the line soup buffer are visualized as green and blue outlines. Segments with a winding direction that points \emph{up} are colored blue while those pointing \emph{down} are colored green. The curves were adapted from the \emph{stills} dataset by \citet{Nehab2020} and the Skia test corpus.}
    \Description{Selection of stroked cubic Béziers exhibiting strong and weak correctness}
    \label{fig:stroke_stills}
\end{figure*}

@@

One of our artifacts is a GPU compute shader that satisfies these criteria. We coupled this with a simple and efficient data layout for parallel processing. Notably, our implementation is fully data-parallel on input path segments and does not require any computationally expensive curve evaluation on the CPU. We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU~[\citenum{WebGPU}]. As such, the ideas presented in this section are portable to various GPU platforms.

@@

An SVG path consists of a sequence of instructions (or ``verbs''). The \emph{lineto} and \emph{curveto} instructions denote an individual \emph{path segment} consisting of either a straight line or a cubic Bézier. The \emph{moveto} instruction is used to begin a new \emph{subpath} and the \emph{closepath} instruction can be used to create a closed contour (by inserting a line connecting the current endpoint to the start of the subpath). A subpath must form a closed contour if painted as a fill. When painted as a stroke, each subpath must begin and end with a cap according to the desired cap style. In addition, each path segment in a stroked subpath must be connected to adjacent path segments of the same subpath with a join. For any given path, the cap and join styles do not vary across subpaths or individual path segments.

@@

\begin{algorithm}
  $tid \gets$ global invocation ID\;
  $(style, segment) \gets readScene(tid)$\;
  \eIf{$style$ is $fill$}{
      \tcp{Output source curve (offset = 0)}
      flatten\_cubic(segment, $0$)\;
  }{
      \tcp{Handle stroked segment}
      \eIf{$segment$ is \emph{stroke cap marker}}{
          \eIf{path is open}{output start cap\;}{output nothing\;}
      }{
          \tcp{Output parallel curves}
          flatten\_cubic(segment,  style.offset)\;
          flatten\_cubic(segment, -style.offset)\;

@@

Each thread outputs a polyline that fits a subset of the rendered subpath's outer contour, represented by the path segment (see Figure \ref{fig:stroke_threads}). For a filled primitive, a thread only outputs the flattened approximation of the path segment. For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{half-width}, and b) the joins that connect the path segment to the next adjacent segment, and c) evolute patches in regions of high curvature. We also store metadata in our input encoding that marks the position of a path segment within the containing subpath. We use this information to determine whether to output an end cap instead of a join. We encode one additional segment designated as the \emph{stroke cap marker} for every subpath. The thread assigned to the marker segment only outputs a start cap, which completes the stroke outline.

@@

We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, and two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}. We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}~[\citenum{wgpu}] framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).

@@

    \item As mentioned earlier, we did not have time to sufficiently explore a pipeline architecture for GPU-based dashing. Given its parameterization by arc length, we believe the Euler spiral representation lends itself well to an approach based on parallel prefix sums of segment arc lengths.
\end{itemize}
